2.3 Prob Model for PCR Positivity Estimation

### 2024-04-09 2_scripts/2.3 Prob Model for PCR pos.qmd
Abbott prob models, xyplot

```{r}
library(dplyr)         # Data management
library(ggplot2)       # ggplot2
library(ggpubr)        # ggplot2 QQ plot for normality check
library(zoo)           # year-month format
library(lmtest)        # for LR test, comparing models
#library(reshape2)     # for melt() function
library(readxl)        # for reading .xlsx data
library(nnet)          # for multinomial log regression
```

@@

# The APL data
load(file = "../1_data/private/RFD4682e.RData")
```

@@

```{r}
## Data-04, 2/8 months data PSO, extracted from Y. Nagakama et al.
X4_Nakagama_all <- read_excel("../1_data/private/Prob Model/4-anonymized_rocheabbott.xlsx",
                              col_types = c("numeric", "text", "numeric", 
                                            "numeric", "numeric", "numeric"))

@@

@@

```{r}
dta_all<-dplyr::bind_rows(dta_01s, dta_03s, 
                          select(dta_04, -c("month_pso")), dta_06s)

@@

@@

# remove all intermediate data
rm(dta_01, dta_01s, dta_03, dta_03s, dta_04, dta_06, dta_06s)
```

@@

#### 1.1 Multinomial Regression Model:

@@

@@

#### 2.2 Apply Model to CBS Data

@@

#### Make by Region Plots

@@

```{r}
# create by-month estimates, by region
# create by-month est for each region
tb_all<-CBSs %>%
  group_by(samplemonth, region) %>%
  summarise(n = n() , 
            ever_inf= mean(ever_inf, na.rm = T), 
            less_3_month=mean(less_3_month, na.rm = T),
            less_6_month=mean(less_6_month, na.rm=T),
            less_9_month=mean(less_9_month, na.rm=T),
            )
# clean NAs
tb_all<-tb_all[!is.na(tb_all$region), ]
```

@@

#### Make plots by region

```{r}
summary(RFD4682_e$N)
# The codebook does not specify what types of anti-N assay is applied, given the number we assume they are using Abbott anti-N assay (small numbers)
```

>>>>

```{r}
library(dplyr)         # Data management
library(ggplot2)       # ggplot2
library(ggpubr)        # ggplot2 QQ plot for normality check
library(zoo)           # year-month format
library(lmtest)        # for LR test, comparing models
#library(reshape2)     # for melt() function
library(readxl)        # for reading .xlsx data
library(nnet)          # for multinomial log regression
#library(expss)         # for labeling variables
library(ggExtra)       # for marginal dist on ggplot
```

@@

# The APL data
load(file = "../1_data/private/RFD4682e.RData")

# The cleaned ver of Canpath data

```

@@

```{r}
## Data-04, 2/8 months data PSO, extracted from Y. Nagakama et al.
## The Convalescent pts data
X4_Nakagama_all <- read_excel("../1_data/private/Prob Model/4-anonymized_rocheabbott.xlsx",
                              col_types = c("numeric", "text", "numeric", 
                                            "numeric", "numeric", "numeric"))

@@

```{r}
## The negative control data, updated on 7-Apr-2024
dta_04_neg <- read_excel("../1_data/private/Prob Model/4-anonymized_rocheabbott.xlsx",
                              sheet = "true_neg",
                              col_types = c("text", "numeric", "numeric", 
                                            "numeric", "numeric", "numeric"))

# correct col names
colnames(dta_04_neg)<-c("gender","age","month_pso","abbott_n","roche_n","pid")
dta_04_neg$pso <- dta_04_neg$month_pso*30

# All participants been sampled only once, wt=1
dta_04_neg$sample_wt<-rep(1, length(dta_04_neg$pid))
dta_04_neg$study<-rep(4.1, length(dta_04_neg$pid))
```

@@

```{r}
dta_all<-dplyr::bind_rows(dta_01s, dta_03s, 
                          select(dta_04, -c("month_pso")), dta_06s)
dta_all<-dplyr::bind_rows(dta_all, 
                          select(dta_04_neg, -c("month_pso")))

@@

# Label the study var
dta_all$study<-as.character(dta_all$study)
dta_all$study[dta_all$study=='1']<-"Leon_FR"
dta_all$study[dta_all$study=='3']<-"Jamaica"
dta_all$study[dta_all$study=='4']<-"Osaka_JP_pos"
dta_all$study[dta_all$study=='4.1']<-"Osaka_JP_neg"
dta_all$study[dta_all$study=='6']<-"UK"

dta_all$time_pso[is.na(dta_all$time_pso)]<-"never_inf"

@@

# remove all intermediate data
rm(dta_01, dta_01s, dta_03, dta_03s, dta_04, dta_04_neg, dta_06, dta_06s)
```

@@

#### 1.1 Multinomial Regression Model (Roche):

@@

#### 1.2 Model mnlm04 same model but for abbott
```{r}
# Model 04  takes sample weights into account
mnlm04<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ log_abbott_n, 
                 data = dta_all, weights = sample_wt)
summary(mnlm04)
## use estimates/SE and 1.96 to calculate p-value manually, should be <0.001
```

@@

#### 2.2 Apply Model (Roche) to CBS Data

@@

#### 2.2.2 Make by Region Plots

@@

```{r}
# create by-month estimates, by region
# create by-month est for each region
tb_all<-CBSs %>%
  group_by(samplemonth, region) %>%
  summarise(n = n() , 
            ever_inf= mean(ever_inf, na.rm = T), 
            less_3_month=mean(less_3_month, na.rm = T),
            less_6_month=mean(less_6_month, na.rm=T),
            less_9_month=mean(less_9_month, na.rm=T),
            )
# clean NAs
tb_all<-tb_all[!is.na(tb_all$region), ]
```
#### 2.2.3 Make plots by region

#### 2.3.1 Load & Clean
```{r}
summary(RFD4682_e$N)
# The codebook does not specify what types of anti-N assay is applied, given the value we assume they are using Abbott anti-N assay (small numbers)

# create a subset of APL
APLs<-RFD4682_e %>%
  select(order_ID, AGE_AT_COLLECTION, GENDER, N, Col_Month, PAT3)

# unify colnames
colnames(APLs)<-c("order_ID","age","gender","abbott_n","samplemonth","fsa")
APLs$samplemonth<-as.yearmon(APLs$samplemonth)

# add log_abbott_n
## find 0 values bf transformation
APLs$abbott_n[APLs$abbott_n==0 & 
              !is.na(APLs$abbott_n)]<-0.001

APLs$log_abbott_n <- log(APLs$abbott_n)
```

#### 2.3.2 Apply model (Abbott)
```{r}
# Apply the model mnlm04
APLs_prob_r<-as.data.frame(predict(mnlm04, newdata = APLs, "probs")) # num of >3 months = 0
mean(APLs_prob_r[,1], na.rm = T)  # never: 53.28%
mean(APLs_prob_r[,4], na.rm = T)  # <3 months: 24.21%
mean(APLs_prob_r[,2], na.rm = T)  # 3-6 month: 5.40%
mean(APLs_prob_r[,3], na.rm = T)  # 6-9 months: 4.66%
mean(APLs_prob_r[,5], na.rm = T)  # >9 months: 12.46%

APLs_prob_r$ever_inf <- 1-APLs_prob_r$never_inf
APLs_prob_r$less_6_month <- APLs_prob_r$less_3_month + APLs_prob_r$`3-6_month`
APLs_prob_r$less_9_month <- APLs_prob_r$less_6_month + APLs_prob_r$`6-9_month`

APLs<-cbind.data.frame(APLs, APLs_prob_r[,c(4,6,7,8)])
```

```{r}
# create melt data (monthly rolling inc prop estimates)
tb_APL<-APLs %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            ever_inf= mean(ever_inf, na.rm = T), 
            less_3_month=mean(less_3_month, na.rm = T),
            less_6_month=mean(less_6_month, na.rm=T),
            less_9_month=mean(less_9_month, na.rm=T),
            )
# clean NAs
tb_APL<-tb_APL[!is.na(tb_APL$samplemonth) & !is.na(tb_APL$ever_inf), ]

```

#### 2.3.3 making Plots
```{r}
tb_melt_APL<-reshape2::melt(select(tb_APL, 1,3:6), 
                        id.vars = c("samplemonth"),
                        variable.name = "inf",
                        value.name = "prop")

p06<-tb_melt_APL %>%
  ggplot(aes(x=samplemonth, y=prop, group=inf, colour=inf)) +
  geom_point(aes(colour=inf)) +
  geom_line(aes(colour=inf))  +
  theme_bw()
p06
```
**Additional plotting** 
```{r}
# merge monthly summary table: CBS + APL
tb_CBS<-tb_all[tb_all$region=="Alberta",] %>%
  select(-2)
tb_CBS$data<-rep("CBS", length(tb_CBS$samplemonth))

tb_APL$data<-rep("APL", length(tb_APL$samplemonth))
tb_APLCBS<-bind_rows(select(tb_CBS, -2), select(tb_APL, -2))

```

```{r}
# create melt data:
tb_melt_2<-reshape2::melt(tb_APLCBS, 
                        id.vars = c("samplemonth", "data"),
                        variable.name = "inf",
                        value.name = "prop")

p07<-tb_melt_2 %>%
  ggplot(aes(x=samplemonth, y=prop, group=data, colour=data)) +
  geom_point(aes(colour=data)) +
  geom_line(aes(colour=data))  +
  theme_bw()
p07 + facet_wrap(~inf)
```

@@

#### 3.4 Plot the correlation btw Roche-N and Abbott-N:
```{r}
# The clear linear relation btw log_roche and log_abbott:
plot(CBSs$log_roche_n, CBSs$log_abbott_n) # this linear association is intriduced by lm01

# How it looks like in normal scale?
plot(CBSs$roche_n, CBSs$abbott_n_hat)     # a smooth curve

dta_all$study<-as.factor(dta_all$study)

# using our training data:
p04<-dta_all %>%
  .[!is.na(.$log_abbott_n) & !is.na(.$log_roche_n),] %>%
  ggplot(aes(x=log_roche_n, y=log_abbott_n, 
             study, colour=study)) +
  geom_point(aes(colour=as.factor(study))) +
  geom_vline(xintercept = 0, linewidth=1.0, linetype="dotted", col="red") +
  geom_hline(yintercept = log(1.4), linewidth=1.0, linetype="dotted", col="red") +
  xlab("Roche Anti-N S/Co ratio (log scale)") + 
  ylab("Abbott Anti-N S/Co ratio (log scale)") +
  labs(color='') +
  scale_colour_manual(labels = c("Jamaica (-)", "Osaka, JP (-)",
                                 "Osaka, JP (+)","UK (+)"), 
                      values = c("red", "skyblue", "darkgreen", "purple")) +
  theme_bw() + 
  theme(legend.position = "bottom")

p04
```

```{r}
## Another figure: add CBS matched data
dta_CBS<-CBSs[!is.na(CBSs$abbott_n) & !is.na(CBSs$roche_n),]  %>%
  select(log_roche_n, abbott_n)
dta_CBS$study<-rep("CBS matched", length(dta_CBS$log_roche_n))
dta_CBS$log_abbott_n<-log(dta_CBS$abbott_n)

dta_CBS<-bind_rows(select(dta_all, study, log_roche_n, log_abbott_n), 
                   select(dta_CBS, study, log_roche_n, log_abbott_n))

p05<-dta_CBS %>%
  .[!is.na(.$log_abbott_n) & !is.na(.$log_roche_n),] %>%
  ggplot(aes(x=log_roche_n, y=log_abbott_n, 
             study, colour=study)) +
  geom_point(aes(colour=as.factor(study))) +
  geom_vline(xintercept = 0, linewidth=1.0, linetype="dotted", col="red") +
  geom_hline(yintercept = log(1.4), linewidth=1.0, linetype="dotted", col="red") +
  xlab("Roche Anti-N S/Co ratio (log scale)") + 
  ylab("Abbott Anti-N S/Co ratio (log scale)") +
  labs(color='') +
  scale_colour_manual(labels = c("CBS","Jamaica (-)", "Osaka, JP (-)",
                                 "Osaka, JP (+)","UK (+)"), 
                      values = c("lightblue","red", "blue", "darkgreen", "purple")) +
  theme_bw() + 
  theme(legend.position = "bottom")

p05
```
#### Alton suggests using marginal dist with {ggExtra}
```{r}
ggMarginal(p04, groupColour = TRUE, groupFill = TRUE) # log-scale
ggMarginal(p05, groupColour = TRUE, groupFill = TRUE) # with CBS matched data
```
